Constraining Quiescent Supermassive Black Hole Environments.

### 2014-08-11 ms.tex
Typo
Rewrite of heating section and cooling

\newcommand{\eddr}{\dot{M}/\dot{M}_{\rm Edd}}
\newcommand{\Mdotb}{\dot{M}_{\rm bondi}}

@@

\newcommand{\dxdy}[2]{\frac{\partial #1}{\partial #2} }
\newcommand{\ddr}[1]{\dxdy{#1}{r}}
\newcommand{\drhodt}{\dxdy{\rho}{t}}
\newcommand{\dpdr}{\dxdy{p}{r}}
\newcommand{\dvdr}{\dxdy{v}{r}}
\newcommand{\dsdr}{\dxdy{s}{r}}
\newcommand{\dphidr}{\dxdy{\Phi}{r}}
\newcommand{\ke}{\frac{v^2}{2}}
\newcommand{\kew}{\frac{v_w^2}{2}}
\newcommand{\gammaf}{\frac{\gamma}{\gamma-1}}
\newcommand{\gammafi}{\frac{\gamma-1}{\gamma}}
\newcommand{\cs}{\frac{p}{\rho}}
\newcommand{\Q}{q (\ke+\kew-\gammaf \cs)}
\newcommand{\kb}{k_{\rm b}}
\renewcommand{\mp}{m_{\rm p}}
\newcommand{\Menc}{M_{\rm enc}}
\newcommand{\Mstar}{M_{*}}
\newcommand{\Mbh}[1][]{M_{\bullet#1}}
\newcommand{\phirs}{\frac{G \Menc}{\rs}}
\newcommand{\soi}{\rm soi}
\newcommand{\ff}{\rm ff}
\newcommand{\rsoi}{r_{\soi}}
\newcommand{\sigsoi}{\sigma_{\soi}}
\newcommand{\vwO}{v_{w,0}}
\newcommand{\x}{\frac{r_s}{\rsoi}}
\newcommand{\vwNorm}{\frac{\vwO}{\sigsoi}}
\newcommand{\pyear}{{\rm yr}^{-1}}

@@

\begin{document}
\title{Constraining Supermassive Black Hole Environments.}
 \maketitle

@@

The gas environment also plays an important role in determining observables (light-curves, SEDs) from jetted tidal disruption events (TDEs). These are bright flares that occur when a star passes within the tidal disruption radius of the SMBH. A small fraction of such events are accompanied by the launch of a relativistic jet into the circumnuclear medium (CNM) of the host galaxy.  As the jet propagates into the CNM, a reverse shock propagates towards the back of the jet, decelerating it. The CNM density profile will determine the final Lorentz factor of the jet. This in turn will strongly affect the duration of the observed synchotron radiation.

@@

We also explore the effects of varying the heating rate on the flow structure. Typically an inflow outflow flow structure is established. The stellar ejecta inside of the stagnation radius ($\rs$) will be accreted, while that outside $\rs$ will be driven out with the outflow. The location of $\rs$ and hence the accretion rate is  quite sensitive to the assumed heating rate.  In fact, as the specific thermal energy of the ejects drops below the potential, an inflow is established out to the radius where the SNe Ia rate becomes comparable to the dynamical timescale, which for black holes smaller than 10$^8 \Msun$ is well outside the $\rsoi$ 
%It turns out that in our models the mass accretion rate is coincidentally quite similar to the Bondi accretion rate.

@@

\begin{align}
\rs=\frac{7}{2}\frac{G \Mbh}{\vw^2}.
\end{align}

@@

%%AG-I still believe that one will not get an outflow without extra heating, but this statement apparently depends on the slope of the density profile...Perhaps some sort of stability arguments could be applied...perturbing density at constant pressure. What happens to the equation for HSE.
%%AG-I believe that in the case of core galaxy, the negative root of the quadratic should be taken...

@@

\section{Discussion}
\subsection{Heating}
We now discuss potential sources of heating and discuss how different physical sources of heating would lead to different values of $\vwO$. For each heating source  from the specific heating rate $E$ we may estimate the corresponding $\vwO$...

@@

Although red giant winds dominate the mass budget, winds from main sequence stars may make a significant contribution to the energy budget, with wind velocities $\sim 100-150$ km/s \citep{NaimanSoares-Furtado+:2013a}. 
%%Even assuming efficient thermalization wouldn't the contribution of main sequence winds be diluted by the fraction they take up of the mass budget.

\item \emph{Type Ia SNe} Following \citealt{ShcherbakovWong+:2014a}, the energy injected per Ia SNe is 10$^{51}$ ergs, and the SNe rate is $\simeq 4 \times 10^{-14} \Msun^{-1}$ yr$^{-1}$.  Assuming that each SNe in 10$^{51}$ ergs and a mass injection rate (from stellar winds) of 10$^{-11} \Msun$  yr$^{-1}$, this leads to $\vwO\simeq 600$ km/s.  (Note one of our chosen values for $\vwO$ is 500 km/s).

As noted in \citealt{ShcherbakovWong+:2014a}, it is not obvious whether or not the SNe energy injection treated as a constant throughout time and space, since the flow time-scale may be short compared to the time between successive SNe Ia...
%AG-Thornton and http://arxiv.org/pdf/1402.6695v2.pdf...
%AG-Case of no stagnation radius
\item \emph{MS Pulsars} 
\end{enumerate}

\subsection{Mass accretion rate}
For each of our solutions, we may calculate the accretion rate $\dot{M}$ onto the SMBH. We show $\Medd$ vs. $\Mbh$ as a function of black hole mass for our three  chosen values of the $\vwO$ (200 km/s, 500 km/s, and 1000 km/s). The results are shown in Figure~\ref{fig:mdot_mass}.  Note that $\dot{M}\sim\vwO^{-2}$ for a cusp galaxy and $\dot{M}\sim\vwO^{-4}$ for a core galaxy. Thus, the mass accretion rate will be a sensitive function of the assumed heating rate.
%%AG-Core galaxies might be a better choice to show the variations in the mass accretion rate 
%%AG-Also want to go into the case of no stagnation radius

@@

%%AG: Split galaxies into cores (squares) and cusps (triangles). For cores only did those galaxies which had all three values of the wind velocity completed. Also want to include a plot of x-ray luminosity (less abstract than the mass accretion rate).

@@

\subsection{Cooling}
\label{sec:cooling}

>>>>

\newcommand{\eddr}{\dot{M}/\dot{M}_{\rm Edd}}
\newcommand{\Mdote}{\dot{M}_{\rm Edd}}
\newcommand{\Mdotb}{\dot{M}_{\rm bondi}}

@@

\newcommand{\dxdy}[2]{\frac{\partial #1}{\partial #2} }
\newcommand{\ddr}[1]{\dxdy{#1}{r}}
\newcommand{\drhodt}{\dxdy{\rho}{t}}
\newcommand{\dpdr}{\dxdy{p}{r}}
\newcommand{\dvdr}{\dxdy{v}{r}}
\newcommand{\dsdr}{\dxdy{s}{r}}
\newcommand{\dphidr}{\dxdy{\Phi}{r}}
\newcommand{\ke}{\frac{v^2}{2}}
\newcommand{\kew}{\frac{v_w^2}{2}}
\newcommand{\gammaf}{\frac{\gamma}{\gamma-1}}
\newcommand{\gammafi}{\frac{\gamma-1}{\gamma}}
\newcommand{\cs}{\frac{p}{\rho}}
\newcommand{\Q}{q (\ke+\kew-\gammaf \cs)}
\newcommand{\kb}{k_{\rm b}}
\renewcommand{\mp}{m_{\rm p}}
\newcommand{\Menc}{M_{\rm enc}}
\newcommand{\rhostar}{\rho_*}
\newcommand{\Mstar}{M_{\star}}
\newcommand{\Mbh}[1][]{M_{\bullet#1}}
\newcommand{\phirs}{\frac{G \Menc}{\rs}}
\newcommand{\soi}{\rm soi}
\newcommand{\ff}{\rm ff}
\newcommand{\rsoi}{r_{\soi}}
\newcommand{\rIa}{r_{\rm Ia}}
\newcommand{\sigsoi}{\sigma_{\soi}}
\newcommand{\vwO}{v_{w,0}}
\newcommand{\x}{\frac{r_s}{\rsoi}}
\newcommand{\vwNorm}{\frac{\vwO}{\sigsoi}}
\newcommand{\pyear}{{\rm yr}^{-1}}

@@

\begin{document}
\title{Constraining Quiescent Supermassive Black Hole Environments.}
 \maketitle

@@

The gas environment also plays an important role in determining observables (light-curves, SEDs) from jetted tidal disruption events (TDEs). These are bright flares that occur when a star passes within the tidal disruption radius of the SMBH. A small fraction of   events are accompanied by the launch of a relativistic jet into the circumnuclear medium (CNM) of the host galaxy.  As the jet propagates into the CNM, a reverse shock propagates towards the back of the jet, decelerating it. The CNM density profile will determine the final Lorentz factor of the jet. This in turn will strongly affect the duration of the observed synchotron radiation.

@@

We also explore the effects of varying the heating rate on the flow structure. Typically an inflow outflow flow structure is established. The stellar ejecta inside of the stagnation radius ($\rs$) will be accreted, while that outside $\rs$ will be driven out with the outflow. The location of $\rs$ and hence the accretion rate is  quite sensitive to the assumed heating rate.  
%In fact, as the specific thermal energy of the ejects drops below the potential, an inflow is established out to the radius where the SNe Ia rate becomes comparable to the dynamical timescale, which for black holes smaller than 10$^8 \Msun$ is well outside the $\rsoi$ 
%It turns out that in our models the mass accretion rate is coincidentally quite similar to the Bondi accretion rate.

@@

\begin{align}
\rs=\frac{7}{2}\frac{G \Mbh}{\vw^2}.
\label{eq:rs_simple}
\end{align}

@@

%%AG-Apparently even in the low heating case, it is still possible to an outflow (from the analytic results above the gas density must be quite steep in this case). However, at low heating we will probably become subject to a cooling instablity.
%%AG-I believe that in the case of core galaxy, the negative root of the quadratic should be taken...

@@

\section{Discussion}
\subsection{Heating}
We now discuss potential sources of heating and discuss how different physical sources of heating would lead to different values of $\vwO$. For each heating source  from the heating rate per unit volume $e$ we may estimate the corresponding $\vwO$, using

\begin{align}
&\frac{q \vwO^2}{2}=e\\
&\frac{\eta \rhostar \vwO^2}{2 t_h}=e
\label{eq:vw_eff}
\end{align}

We estimate the value of the $\vwO$ for different physical heating sources below.

@@

Although red giant winds dominate the mass budget, winds from main sequence stars may make a significant contribution to the energy budget, with wind velocities $\sim 100-150$ km/s \citep{NaimanSoares-Furtado+:2013a}.

\item \emph{Type Ia SNe} Following \citealt{ShcherbakovWong+:2014a}, the energy injected per Ia SNe is $E_{\rm Ia}=10^{51}$ ergs, and the SNe rate is $R_{\rm Ia}\simeq 4 \times 10^{-14} \Msun^{-1}$ yr$^{-1}$.

\begin{align}
e=R_{\rm Ia} E_{\rm Ia} \rhostar
\end{align}

Then from Equation~\ref{eq:vw_eff}

\begin{align}
\vwO=\sqrt{\frac{2 t_h R_{\rm Ia} E_{\rm Ia}}{\eta}}
\end{align}

 Following \citealt{ShcherbakovWong+:2014a}, $E_{\rm Ia}=10^{51}$ ergs and $R_{\rm Ia}\simeq 4 \times 10^{-14} \Msun^{-1}$ yr$^{-1}$, which gives $eta\simeq 230 \eta^{-1/2}$ km/s (For $\eta=0.1$ this corresponds to $\vwO\simeq$ 500 km/s.   
 %Perhaps it would be best to adopt a fiducial eta=0.1--this may be more straightforward from the pov of both heating and cooling.

As noted in \citealt{ShcherbakovWong+:2014a}, SNe energy injection cannot necessarily be treated as constant throughout  time and space, since the flow time-scale may be short compared to the time between successive SNe Ia.

There will be some radius, $r_{\rm Ia}$, where the dynamical timescale of the flow becomes comparable to the time between successive Type Ia SNe.  If the stagnation radius of the flow is outside of $\rIa$ then SNes will provide a steady contribution to $\vwO$

The timescale between successive Type Ia SNes  is

\begin{align}
t_{\rm Ia} \sim 2.5 \times 10^5  \Mstar_{,8}^{-1} \; {\rm yrs}.
\end{align}

The dynamical time is given

\begin{align}
 t_{\rm dyn} \sim r/\sigma \sim G M_\star/\sigma^3 \sim 5 \times 10^4 \Mstar_{,8} \sigma_{200}^{-3} \; {\rm yrs}.
\end{align}

Therefore,

\begin{align}
 r_{\rm Ia} \sim G \Mstar_{\rm Ia}/\sigma^2 \sim 4 \sigma_{200}^{-1/2} \; {\rm pc}
\end{align}

%AG-Thornton and http://arxiv.org/pdf/1402.6695v2.pdf...

\item \emph{MS Pulsars} 
The spin down power of millisecond pulsars (MSPs) could potentially be an important heating source. Assuming a spin-down luminosity $L_{\rm sd}$ and $n_{\rm msp}$ MSPs per unit mass

\begin{align}
e=L_{\rm sd} n_{\rm msp} \rhostar.
\end{align}

Then we may obtain the $\vwO$ from Equation~\ref{eq:vw_eff}. m

\begin{align}
\vwO=\sqrt{\frac{2 t_h n_{\rm msp} L_{\rm sd}}{\eta}}
\end{align}

There may be as many 30,000 MSPs in the Milky way, which would give  $n_{\rm msp}\simeq3 \times 10^{-41} $ MSPs g$^{-1}$. Even if we optimistically take $L_{\rm sd}\simeq 10^{35}$ ergs/s, and $\eta$=0.1, we would get $\vwO\simeq40$ km/s, which means heating by MSPs is negligible.

Note MSPs may also be dissociated in the galactic center...
%%Also low thermalization efficiency inferred in 47 Tuc by Naiman et al.
\end{enumerate}

\subsection{Mass accretion rate}
For each of our solutions, we may calculate the accretion rate $\dot{M}$ onto the SMBH. We show $\Mdote$ vs. $\Mbh$ as a function of black hole mass for our three  chosen values of the $\vwO$ (200 km/s, 500 km/s, and 1000 km/s). The results are shown in Figure~\ref{fig:mdot_mass}.  Note that $\dot{M}\sim\vwO^{-2}$ for a cusp galaxy and $\dot{M}\sim\vwO^{-4}$ for a core galaxy. Thus, the mass accretion rate will be a sensitive function of the assumed heating rate.
%%AG-The variation in accretion rate for w/ vw for core galaxies is comparable to that of cusp galaxies and belies the steep dependence stated above.
%%AG-Note separation of cores and cusps (at vw=1000 km/s, where there are enough core galaxies for this comparison to be made). Also, note that the spread in core galaxies is far larger.

The accretion rate is proportional to the stellar mass enclosed inside of $\rs$, so for core galaxies $\dot{M}\sim \Mbh (\rs/\rsoi)^{2}$, while for cusp galaxies $dot{M}\sim \Mbh (\rs/\rsoi)$.  This steeper dependence is enough to offset the tendency for core galaxies to have a larger $\rs/\rsoi$ for fixed $\eta$ (see Figure~\ref{fig:stag})Therefore, for $\rs/\rsoi<1$ core galaxies would have a smaller $\dot{M}$ than cusp galaxies.  For $\vwO=1000$ km/s, core galaxies have a smaller accretion rate for a given black hole mass.

@@

%%AG:  Also want to include a plot of x-ray luminosity (less abstract than the accretion rate).

@@

\subsection{Cooling}
\label{sec:cooling}
In our models stellar winds collide, shock heat, and thermalize. The rate of energy injection (parameterized by $\vwO$) sets a characteristic temperature--see Equation~\ref{eq:tstag}.  For a given amount of injected mass, the energy input from shock heating should be equal to the enthalpy of the gas.

In principle, the gas should radiatively cool as well--either from metal line cooling for $T\lsim 2\times 10^7$ K, or through free-free emission for $T\gsim 2\times 10^7$ K.  If the cooling rate is much smaller than the rate of energy injection then radiative cooling may be safely neglected.

For each of our solutions we compare the radiative cooling rate, $C$ to the heating rate $H=q \vw^2/2$. If $H/C \gg  1$ then cooling may be neglected. We take

\begin{align}
c=\Lambda(T) n^2.
\end{align}

Where $n$ is the number density and $\Lambda(T)$ is the cooling function. We approximate $\Lambda(T)$ as piece-wise power law (fitted for solar metallicity--see Figure 34.1 in \cite{Draine:2011a})
%cite Draine

\[
\Lambda(T)\simeq
\begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}& \text{if } T \gsim 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}}& \text{if } T \lsim 2 \times 10^7 \text{K}     
\end{cases}
\label{eq:cooling}
\]

Note $C\propto n^2\propto \eta^2$. On the other hand, $H \propto q  \propto \eta$. Thus, the ratio $H/C$ will depend upon the free parameter $\eta$. We plot $H/C$ for all of our profiles for our fiducial $\eta=1$.

\begin{figure}
\includegraphics[width=\columnwidth]{cooling.eps}
\caption{\label{fig:cooling} Ratio of heating ($H$) to cooling rate ($C$) for each of our solutions. The cooling rate is calculated using the approximate (solar metallicity) cooling function in Equation~\ref{eq:cooling}. The heating rate is estimated as $q \vw^2/2$. The ratio is computed assuming $eta=1$.}
\end{figure}

For $\vwO$=1000 km/s, it is apparent from Figure~\ref{fig:cooling}, that radiative cooling may be safely neglected. With a few exceptions this is also true for $\vwO$=500 km/s.  However, for $\vwO$=200 km/s cooling may be quite important. However, at smaller $\eta$ the effects of cooling will be less pronounced. For each of our galaxies we give the maximum $\eta$ such that $H/C >$1 in Table .  Roughly speaking, this is the maximum $\eta$ for which it would be safe to ignore cooling. $\eta$ depends on the age of the stellar population, and will realistically vary between $\sim 0.1$ and $\sim 1$.
